---
title: 'Assignment 9: Spatial Analysis in R'
author: "Andreana Chou"
output:
  html_document: default
  pdf_document: default
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics (ENV872L) on spatial analysis.

## Directions
1. Rename this file `<FirstLast>_A09_SpatialAnalysis.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Use the lesson as a guide. It contains code that can be modified to complete the assignment.
4. Work through the steps, **creating code and output** that fulfill each instruction.
5. Be sure to **answer the questions** in this assignment document. Space for your answers is provided in this document and is indicated by the ">" character. If you need a second paragraph be sure to start the first line with ">". You should notice that the answer is highlighted in green by RStudio.
5. When you have completed the assignment, **Knit** the text and code into a single HTML file.


## DATA WRANGLING

### Set up your session

1. Import libraries: tidyverse, sf, leaflet, here, and mapview
2. Execute the `here()` command to display the current project directory

```{r Prepare workspace, error=F, warning=F}
#1. 

#import libraries
library(tidyverse)
library(sf)
library(leaflet)
library(here)
library(mapview)

#2. 

#display here()
here()

```

### Read (and filter) county features into an sf dataframe and plot

In this exercise, we will be exploring stream gage height data in Nebraska corresponding to floods occurring there in 2019. First, we will import from the US Counties shapefile we've used in lab lessons, filtering it this time for just Nebraska counties. Nebraska's state FIPS code is `31` (as North Carolina's was `37`).

3. Read the `cb_2018_us_county_20m.shp` shapefile into an sf dataframe, filtering records for Nebraska counties (State FIPS = 31)
4. Reveal the dataset's coordinate reference system
5. Plot the records as a map (using `mapview` or `ggplot`)

```{r Read the county data into an sf dataframe}
#3. Read in Counties shapefile into an sf dataframe, filtering for just NE counties

#used st_read() with here() functions, filtered by STATEFP column
Nebraska.counties <- st_read(here("./Data/Spatial/cb_2018_us_county_20m.shp")) %>%
  filter(STATEFP == 31)

#4. Reveal the CRS of the counties features

st_crs(Nebraska.counties)

#5. Plot the data

mapView(Nebraska.counties)

```

6. What is the EPSG code of the Counties dataset? Is this a geographic or a projected coordinate reference system? (Or, does this CRS use angular or planar coordinate units?) To what datum is this CRS associated? (Tip: look for the EPSG code on <https://spatialreference.org> and examine the link for "Well Known Text as HTML" or "Human-Readable OGC WKT"...)

> ANSWER: The EPSG code for the Counties dataset is 4269. This is associated with the North America datum, and is a geographic coordinate system that records location using angular units. 
> 

### Read in gage locations csv as a dataframe, then display the column names it contains

Next we'll read in some USGS/NWIS gage location data added to the `Data/Raw` folder. These are in the `NWIS_SiteInfo_NE_RAW.csv` file.(See `NWIS_SiteInfo_NE_RAW.README.txt` for more info on this dataset.)

7. Read the NWIS_SiteInfo_NE_RAW.csv file into a standard dataframe, being sure to set the `site_no` field as well as other character columns as a factor. 

8. Display the column names of this dataset.

```{r Read in USGS Gage Locations}
#7. Read in gage locations csv as a dataframe

#used read.csv to ensure data is read as a dataframe
Nebraska.NWIS <- read.csv(here("./Data/Raw/NWIS_SiteInfo_NE_RAW.csv"), stringsAsFactors = TRUE)

#converted site_no to factor data type with as.factor()
Nebraska.NWIS$site_no <- as.factor(Nebraska.NWIS$site_no)

#8. Reveal the names of the columns

colnames(Nebraska.NWIS)

```

9. What columns in the dataset contain the x and y coordinate values, respectively?  
> ANSWER: dec_long_va (longitude) is the x-coordinate value, dec_lat_va (latitude) is the y-coordinate value.
> 


### Convert the dataframe to a spatial features ("sf") dataframe

10. Convert the dataframe to an sf dataframe.
 * Note: These data use the same coordinate reference system as the counties dataset

11. Display the column names of the resulting sf dataframe

```{r Convert dataframe to sf dataframe}
#10. Convert to an sf object

#used st_as_sf() function and established dec_long_va/dec_lat_va as the 
#geometric coordinates, with the same crs as counties dataset
Nebraska.NWIS.sf <- Nebraska.NWIS %>% st_as_sf(coords=c("dec_long_va","dec_lat_va"),
                                               crs=4269)

#11. Re-examine the column names

colnames(Nebraska.NWIS.sf)
```

12. What new field(s) appear in the sf dataframe created? What field(s), if any, disappeared?

>ANSWER: The dec_long_va and dec_lat_va columns were replaced by the geometry column.
>


### Plot the gage locations on top of the counties

13. Use `ggplot` to plot the county and gage location datasets.
 * Be sure the datasets are displayed in different colors
 * Title your plot "NWIS Gage Locations in Nebraska"
 * Subtitle your plot with your name

```{r Plot the spatial features}
#13. Plot the gage locations atop the county features

#used ggplot() and geom_sf() to layer gage location over counties, used labs()
#to add titles/subtitles
#used color=red instead of fill=red to ensure all data points were red, 
#instead of black

NWIS.Nebraska.plot <- ggplot() + 
  geom_sf(data=Nebraska.counties, fill="lightblue") +
  geom_sf(data=Nebraska.NWIS.sf, color="red") +
  labs(title = "NWIS Gage Locations in Nebraska", 
       subtitle = "Andreana Chou")
NWIS.Nebraska.plot

```


### Read in the gage height data and join the site location data to it.

Lastly, we want to attach some gage height data to our site locations. I've constructed a csv file listing many of the Nebraska gage sites, by station name and site number along with stream gage heights (in meters) recorded during the recent flood event. This file is titled `NWIS_SiteFlowData_NE_RAW.csv` and is found in the Data/Raw folder.

14. Read the `NWIS_SiteFlowData_NE_RAW.csv` dataset in as a dataframe
 * Pay attention to which fields should be imported as factors!
15. Show the column names .
16. Join our site information (already imported above) to these gage height data
 * The `site_no` and `station_nm` can both/either serve as joining attributes
 * Construct this join so that the result only includes spatial features where both tables have data
17. Show the column names in this resulting spatial features object
18. Show the dimensions of the resulting joined dataframe

```{r Read in gage site data}
#14. Read the site flow data into a data frame

#read.csv() function
NWIS.SiteFlow <- read.csv(here("./Data/Raw/NWIS_SiteFlowData_NE_RAW.csv"), 
                          stringsAsFactors = TRUE)

#set date as date type with as.Date(), site_no as factor with as.factor()
NWIS.SiteFlow$date <- as.Date(NWIS.SiteFlow$date)
NWIS.SiteFlow$site_no <- as.factor(NWIS.SiteFlow$site_no)

#15. Show the column names

colnames(NWIS.SiteFlow)

#16. Join the flow data to our NWIS gage location spatial dataframe

#used left_join() to attach site flow data with original data frame (Nebraska.NWIS.sf)
#used drop_na() on the gage_ht column to eliminate N/As

NWIS.joined.sf <- Nebraska.NWIS.sf %>% left_join(NWIS.SiteFlow, by="site_no") %>% drop_na(gage_ht)

#17. Show the column names of the joined dataset

colnames(NWIS.joined.sf)

#18. Show the dimensions of this joined dataset

dim(NWIS.joined.sf)

```

### Map the pattern of gage height data

Now we can examine where the flooding appears most acute by visualizing gage heights spatially.
19. Plot the gage sites on top of counties (using `mapview`, `ggplot`, or `leaflet`) 
 * Show the magnitude of gage height by color, shape, other visualization technique.

```{r map gage sites}
#Map the points, sized by gage height

#combined two mapView() functions, with cex="gage_ht" to show the magnitude of gage
#height via circle size on NWIS.joined.sf's mapView()
Nebraska.gageheight <- mapView(Nebraska.counties, col.regions="lightblue") +
  mapView(NWIS.joined.sf, col.regions="purple", cex="gage_ht")

Nebraska.gageheight

```

---

## SPATIAL ANALYSIS

Up next we will do some spatial analysis with our data. To prepare for this, we should transform our data into a projected coordinate system. We'll choose UTM Zone 14N (EPGS = 32614).


### Transform the counties and gage site datasets to UTM Zone 14N

20. Transform the counties and gage sf datasets to UTM Zone 14N (EPGS = 32614).
21. Using `mapview` or `ggplot`, plot the data so that each can be seen as different colors

```{r spatial analysis}
#20 Transform the counties and gage location datasets to UTM Zone 14

#used st_transform() with sf datasets and EPGS code 32614 as arguments

NE.counties.utm <- st_transform(Nebraska.counties, 32614)
NE.gage.utm <- st_transform(Nebraska.NWIS.sf, 32614)

#21 Plot the data

#combined two mapView() functions, one for each utm respectively

mapView(NE.counties.utm, col.regions="lightblue") + 
  mapView(NE.gage.utm, col.regions="red")

```


### Select the gages falling within a given county

Now let's zoom into a particular county and examine the gages located there. 
22. Select Dawson county from your projected county sf dataframe 
23. Select the gage sites falling `within` that county
 * Use either matrix subsetting or tidy filtering 
24. Create a plot showing:
 * all Nebraska counties,
 * the selected county, 
 * and the gage sites in that county

```{r spatial selection}
#22 Select the county

#filter Nebraska.counties sf for Dawson county under "NAME" column
Dawson.sf <- Nebraska.counties %>% filter(NAME == "Dawson")

#23 Spatially select gages within the selected county

#intersected the two sf objects with matrix method

Dawson.gage.intersect <- NWIS.joined.sf[Dawson.sf, ]

#24 Plot

#combined three mapView() functions, layering Dawson gage sites on top of Dawson
#county outline, on top of all Nebraska county outlines

mapView(Nebraska.counties, col.regions = "lightgreen") + 
  mapView(Dawson.sf, col.regions="lightpink") + 
  mapView(Dawson.gage.intersect, col.regions="darkblue")

```
