---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Andreana Chou"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
knitr::opts_chunk$set(warning = FALSE)
#1

#loaded libraries tidyverse, agricolae, ggplot2, lubridate, here
library(tidyverse)
library(agricolae)
library(ggplot2)
library(lubridate)
library(here)

#imported NTL-LTER chemistry/physics data file
NTL_LTER <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                     stringsAsFactors = TRUE)

#converted to dates with lubridate
NTL_LTER$sampledate <- mdy(NTL_LTER$sampledate)

#2

#assigned custom theme to my_theme object
my_theme <- theme_classic(base_size = 12) + 
  theme(panel.background = element_rect(color = "lightblue", fill = "white"),
        panel.grid.major = element_line(color = "lightblue", linewidth = 0.5),
        legend.title = element_text(color = "black"),
        legend.position = "right")

#set custom theme as default
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: Mean lake temperature recorded during July does not change with depth across all lakes.
Ha: Mean lake temperature recorded during July does differ with depth across all lakes. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

\newpage

```{r scatterplot}
#4

#wrangle data using pipe function, select, filter, and drop_na
NTL_LTER4 <- NTL_LTER %>% 
  select("lakename", "year4", "daynum", "depth", "temperature_C") %>% 
  filter(between(daynum, 182, 213)) %>% 
  drop_na(temperature_C)

#5

#created scatter plot with geom_point, used geom_smooth(method="lm") to create line 
#of best fit, ylim to set upper and lower bounds on y-axis
temp_depth_scatter <- NTL_LTER4 %>% ggplot(aes(x=depth, y=temperature_C)) + 
  geom_point() + 
  geom_smooth(method="lm") + ylim(0, 35) + 
  labs(x="Lake Depth (m)", y="Temperature (C)", 
       title="Peter & Paul Lakes: July Lake Depth Temperatures")
temp_depth_scatter

```

\newpage

6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests temperature will decrease with lake depth, but plateaus
after a depth of 10 meters. The distribution of points suggests a linear trend, but
mathematical analysis is needed to determine how strong the relationship is between
temperature and lake depth. 


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

#used lm to perform linear regression on temperature and lake depth variables
temp_depth_regression <- lm(data=NTL_LTER4, depth ~ temperature_C)
summary(temp_depth_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: 0.7368 (or 73.68%) of the variability in temperature is explained by changes in depth, based on 9974 degrees of freedom. For every meter increase of depth, the temperature decreases by 0.37 degrees C. These findings are statistically significant because the p-value is less than 0.05. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

#assigned multi-variable linear regression model to temp_var AIC object
temp_var <- lm(data = NTL_LTER4, temperature_C ~ year4 + daynum + depth)

#performed AIC using AIC object
step(temp_var)

#10

#performed linear regression on AIC's recommended variables
temp_regression <- lm(data = NTL_LTER4, temperature_C ~ year4 + daynum + depth)
summary(temp_regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables suggested by AIC is year4, daynum, and depth. This model explains 0.7398 (73.98%) of the observed variance in temperature. This is a slight (0.3) improvement over the model that uses only depth as the explanatory variable.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

#wrangle data by parsing lakename and depth categories, used summarise function to 
#create average temperature column
lakes_average <- NTL_LTER4 %>% group_by(lakename, depth) %>% 
  summarise(temp_C_average = mean(temperature_C)) 

#performed ANOVA model between average temperature and lake name
lakes_anova <- aov(data=lakes_average, temp_C_average ~ lakename)
summary(lakes_anova)

#performed linear regression model between temperature and lake name
lakes_anova2 <- lm(data=lakes_average, temp_C_average ~ lakename)
summary(lakes_anova2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is not a significant difference in mean temperatures among the lakes because the p-value from both ANOVA and linear models is 0.0864, which is greater than 0.05. Therefore, the null hypothesis that there is no significant difference in mean temperatures, cannot be rejected.

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.

#created scatterplot with geom_point, alpha=0.5 denotes transparency of points,
#geom_smooth(method="lm") to create line of best fit, ylim to set upper 
#and lower bounds on y-axis

temp_depth_scatter2 <- NTL_LTER4 %>% 
  ggplot(aes(x=depth, y=temperature_C, color=lakename)) + 
  geom_point(alpha=0.5) + 
  geom_smooth(method="lm") + ylim(0, 35) + 
  labs(x="Lake Depth (m)", y="Temperature (C)", 
       title="July Lake Depth Temperatures", color="Lake Name") 
temp_depth_scatter2

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
#Tukey HSD test contains ANOVA model as argument
TukeyHSD(lakes_anova)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Central Long Lake, Crampton Lake, East Long Lake, Hummingbird Lake, Paul Lake, Ward Lake, West Long Lake all have the statistically same mean temperatures as Peter Lake. There are no lakes with statistically distinct mean temperatures from each other.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Two-sample t-test. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

#wrangle data by filtering out Crampton Lake and Ward Lake names
crampton_ward <- NTL_LTER4 %>% filter(lakename == c("Crampton Lake", "Ward Lake"))

#t-test function to analyze temperature by lake name
t.test(crampton_ward$temperature_C ~ crampton_ward$lakename, var.equal = TRUE)
```

>Answer: The mean of Crampton Lake (15.54 degrees C) and Ward Lake (14.25 degrees C) are different. However, the t-value is 1.15 with 236 degrees of freedom and a p-value of 0.2513. With a p-value > 0.05, the null hypothesis that "there is no significant difference in Crampton Lake and Ward Lake July temperatures" remains. This aligns with the answer from number 16. 
