---
title: "Assignment 4: Data Wrangling"
author: "Andreana Chou"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions

1. Rename this file `<FirstLast>_A04_DataWrangling.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.
6. Ensure that code in code chunks does not extend off the page in the PDF.

The completed exercise is due on Thursday, Sept 28th @ 5:00pm.

## Set up your session 

1a. Load the `tidyverse`,  `lubridate`, and `here` packages into your session.

1b. Check your working directory.

1c. Read in all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Apply the `glimpse()` function to reveal the dimensions, column names, and structure of each dataset. 

```{r setup, message = FALSE}
#1a 
library(tidyverse)
library(lubridate)
library(here)

#1b 

print(getwd())     #working directory is correct path

#1c 

O3_18 <- read.csv(here("./Data/Raw/EPAair_O3_NC2018_raw.csv"), stringsAsFactors = TRUE)
O3_19 <- read.csv(here("./Data/Raw/EPAair_O3_NC2019_raw.csv"), stringsAsFactors = TRUE)
PM25_18 <- read.csv(here("./Data/Raw/EPAair_PM25_NC2018_raw.csv"), stringsAsFactors = TRUE)
PM25_19 <- read.csv(here("./Data/Raw/EPAair_PM25_NC2019_raw.csv"), stringsAsFactors = TRUE)

#2 

glimpse(O3_18)

glimpse(O3_19)

glimpse(PM25_18)

glimpse(PM25_19)
```


## Wrangle individual datasets to create processed files.

3. Change the Date columns to be date objects.

4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE

5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).

6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r wrangle data}
#3

O3_18$Date <- mdy(O3_18$Date)     #original dataset had dates set in month-day-year format,
                                  
O3_19$Date <- mdy(O3_19$Date)

PM25_18$Date <- mdy(PM25_18$Date)

PM25_19$Date <- mdy(PM25_19$Date)

#4

#used a pipe function to simplify select() function, assigned to new dataset

O3_18_filter <- O3_18 %>% select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                                 COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

O3_19_filter <- O3_19 %>% select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                                 COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

PM25_18_filter <- PM25_18 %>% select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                                     COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

PM25_19_filter <- PM25_19 %>% select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, 
                                     COUNTY, SITE_LATITUDE, SITE_LONGITUDE)

#5

#used a mutate() function to change all values under an existing column with "="

PM25_18_filter2 <- mutate(PM25_18_filter, AQS_PARAMETER_DESC = "PM2.5")

PM25_19_filter2 <- mutate(PM25_19_filter, AQS_PARAMETER_DESC = "PM2.5")

#6

#used write.csv() function to save and locate processed datasets

write.csv(O3_18_filter, row.names=FALSE, 
          file = here("./Data/Processed/EPAair_O3_NC2018_processed.csv"))

write.csv(O3_19_filter, row.names=FALSE, 
          file = here("./Data/Processed/EPAair_O3_NC2019_processed.csv"))

write.csv(PM25_18_filter2, row.names=FALSE,
          file=here("./Data/Processed/EPAair_PM25_NC2018_processed.csv"))

write.csv(PM25_19_filter2, row.names=FALSE,
          file=here("./Data/Processed/EPAair_PM25_NC2019_processed.csv"))

```


## Combine datasets

7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code.

8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

  * Include only sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels - but it will include sites with missing site information, which you don't want...) 

  * Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site name, AQS parameter, and county. Take the mean of the AQI value, latitude, and longitude.
  
  * Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
  
  * Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.

10. Call up the dimensions of your new tidy dataset.

11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r combining datasets}
#7 

#created new dataset and used rbind() function

combo <- rbind(O3_18_filter, O3_19_filter, PM25_18_filter2, PM25_19_filter2)

#8 

#created a new dataset and used pipe function to simplify wrangling

#used filter() function with "%in% c()" with site names as strings to mass

#filter commonalities across the 4 original datasets

#used group_by() function with columns as arguments

#used summarise() function to re-assign new values (the mean) to existing 
#columns, .groups='drop' was added due to R Studio error suggestions

#used mutate to create a new column by setting it equal to lubridate functions
#month() and year()

combo_wrangle <- combo %>% 
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", 
                          "Hattie Avenue", "Clemmons Middle", "Mendenhall School", 
                          "Frying Pan Mountain", "West Johnston Co.", 
                          "Garinger High School", "Castle Hayne", 
                          "Pitt Agri. Center", "Bryson City", "Millbrook School") ) %>%   
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(DAILY_AQI_VALUE=mean(DAILY_AQI_VALUE), 
            SITE_LATITUDE=mean(SITE_LATITUDE), 
            SITE_LONGITUDE=mean(SITE_LONGITUDE)) %>%
  mutate(month=month(Date)) %>%
  mutate(year=year(Date))

#9

#created a new dataset and used spread() function with arguments: dataset,
#key (column whose values become variable names), value (column whose values
#will be filled under their respective keys)

combo_wrangle2 <- spread(combo_wrangle, AQS_PARAMETER_DESC, DAILY_AQI_VALUE)

#10

#used dim() function

print(dim(combo_wrangle2))

#11

#used write.csv() function to save and locate processed dataset

write.csv(combo_wrangle2, row.names=FALSE,
          file=here("./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv"))

```


## Generate summary tables

12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where mean **ozone** values are not available (use the function `drop_na` in your pipe). It's ok to have missing mean PM2.5 values in this result. 

13. Call up the dimensions of the summary dataset.

```{r summary tables}

#12

#created new dataset and used pipe function to simplify wrangling

combo_wrangle2_sum <- combo_wrangle2 %>%
  group_by(Site.Name, month, year) %>%
  
  #used group_by() to aggregate/organize rows by site name, month, and year
  
  mutate(Mean_AQI_Ozone = mean(Ozone)) %>%
  mutate(Mean_AQI_PM = mean(PM2.5)) %>%
  
  #used mutate() to create a new column that derived values (the mean) from 
  #another column (Ozone and PM2.5)
  
  drop_na(Mean_AQI_Ozone)

  #used drop_na() for mutated ozone column only

#13

#used dim() function

print(dim(combo_wrangle2_sum))

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: na.omit is used for vectors and matrices containing a sequence of values, which means it only removes "NA" from the vectors and matrices. On the other hand, drop_na will get rid of the entire row if any cell within that row has "NA". 
